\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity}

### 2013-12-09 docs/paper/paper.tex
Moar methods!

%Question: should this be included?
\subsubsection*{Functional connectivity}
Functional connectivity describes the statistical dependence of neuronal activity between different brain regions \cite{friston1993functional}.
This gives an insight in the organisation of the brain.
Functional connectivity is strongly time-dependent, as activity changes rapidly providing and fMRI has a low temporal resolution.
This means measured dependencies can be the result of statistical noise.
Functional connectivity can be deduced by fMRI measurement of brains in resting-state \cite{Lowe2000, doria2010, Bullmore2009}.
Resting state indicates that subjects are instructed to relax without thinking of anything in particular to stimulate spontaneous brain activity.

\subsubsection*{Structural connectivity}
Structural connectivity is defined as the mapping of anatomical - neural - paths in the brain between different brain regions \cite{friston1994}.
This is strongly related with functional connectivity: regions can only be functionally connected if there is a structural relation between them \cite{cabral2012}.
This intuitive relation can be demonstrated empirically \cite{vandenheuvel2009}.
Structural connectivity is less time dependent as it involves mappings of anatomical connections opposing temporal activity patterns.

@@

\subsubsection*{Effective connectivity}
In contrast to functional and structural connectivity, effective connectivity takes into account the cause and effect of relations.
It has been described as ``the influence one neural system exerts over another'' \cite{friston1994}.
Effective connectivity indicates which brain regions stimulate other regions.
Time series need to be analysed to infer effective connectivity as cause and effect can be deduced from which event precedes which. 
It is possible to infer effective connectivity from structural and functional connectivity \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.

@@

\section{Methods}
% Different adaptations of PC
The original PC algorithm assumes causal sufficiency, excluding the existence of common confounders between nodes.
This seems unwanted when handling brain data, where one would expect a large amount of latent processes.
Also measuring methods are not yet advanced enough to measure the full detail of the brain, thus skipping over details.
In the perspective of causal inference, these missing areas can also appear as confounders.

@@

\paragraph{EMS-PC}
We have combined the above ideas into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
Two changes have been made in this adaption with respect to modified PC as described by Spirtes et al \cite{Spirtes2000}. In the algorithm as described in section ~\ref{sec:pseudocode}, the search for a separating sets for $X$ and $Y$ stops when such a set is found; instead, EMS continues the search for more separating sets of the same order.

@@

\subsection*{Data acquisition \& analysis}
% TODO: vraag Max naar de details van de meetmethode en subject acquisition
In order 
We will apply the mentioned methods on structural and functional data gathered from human subjects.
For this we are largely dependent on data collected by others.
Hinne et al. has provided fMRI time-series data and structural network data as used and derived in \cite{hinne2013}, of which six subjects were selected at random.
We can estimate functional connectivity from fMRI-measurement in combination with structural 
networks according to a procedure currently submitted by Hinne et al. \cite{hinne2013submitted}.
To find effective connectivity, both structural networks and functional data is needed.

>>>>

%Question: should this be included?
\subsubsection{Functional connectivity}
Functional connectivity describes the statistical dependence of neuronal activity between different brain regions \cite{friston1993functional}.
This gives an insight in the organisation of the brain.
Functional connectivity is strongly time-dependent, as activity changes rapidly providing and fMRI has a low temporal resolution.
This means measured dependencies can be the result of statistical noise.
Functional connectivity can be deduced by fMRI measurement of brains in resting-state \cite{Lowe2000, doria2010, Bullmore2009}.
Resting state indicates that subjects are instructed to relax without thinking of anything in particular to stimulate spontaneous brain activity.

\subsubsection{Structural connectivity}
Structural connectivity is defined as the mapping of anatomical - neural - paths in the brain between different brain regions \cite{friston1994}.
This is strongly related with functional connectivity: regions can only be functionally connected if there is a structural relation between them \cite{cabral2012}.
This intuitive relation can be demonstrated empirically \cite{vandenheuvel2009}.
Structural connectivity is less time dependent as it involves mappings of anatomical connections opposing temporal activity patterns.

@@

\subsubsection{Effective connectivity}
In contrast to functional and structural connectivity, effective connectivity takes into account the cause and effect of relations.
It has been described as ``the influence one neural system exerts over another'' \cite{friston1994}.
Effective connectivity indicates which brain regions stimulate other regions.
Time series need to be analysed to infer effective connectivity as cause and effect can be deduced from which event precedes which. 
It is possible to infer effective connectivity from structural and functional connectivity \cite{mclntosh1994, harrison2003, friston2003, roebroeck2005}.

@@

\section{Material and Methods}
% Different adaptations of PC
The original PC algorithm assumes causal sufficiency, excluding the existence of common confounders between nodes.
This seems unwanted when handling brain data, where one would expect a large amount of latent processes.
Also measuring methods are not yet advanced enough to measure the full detail of the brain, thus skipping over details.
In the perspective of causal inference, these missing areas can also appear as confounders.

@@

\paragraph{EMS-PC}
We have combined the above ideas into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
Two changes have been made in this adaption with respect to modified PC as described by Spirtes et al \cite{Spirtes2000}.
% TODO: line below is repetition of previous paragraph. Keep?
In the algorithm as described in section ~\ref{sec:pseudocode}, the search for a separating sets for $X$ and $Y$ stops when such a set is found; instead, EMS continues the search for more separating sets of the same order.

@@

\subsection{Experimental validation}
% Section on how we plan on demonstrating the empirical value of research.
% Full description from step 1 to final data set
Our analysis consisted of two parts, corresponding to the two different parts of PC algorithm.
Firstly we discerned structure by using brain resting-state time series fMRI data with structure PC.
The generated structure is the same for all variants of PC discussed and proposed, as only the first separating set is used in finding unshielded triples.
We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging.

\subsection{Data acquisition}
% TODO: vraag Max naar de details van de meetmethode en subject acquisition
The above analysis was conducted on the same dataset as in \cite{hinne2013structfunc}. %TODO check citation.
The dataset consists of fMRI

We will apply the mentioned methods on structural and functional data gathered from human subjects.
For this we are largely dependent on data collected by others.
Hinne et al. has provided fMRI time-series data and structural network data as used and derived in \cite{hinne2013}, of which six subjects were selected at random.
We can estimate functional connectivity from fMRI-measurement in combination with structural 
networks according to a procedure by Hinne et al. \cite{hinne2013structfunc}.
To find effective connectivity, both structural networks and functional data is needed.
